---
title: "Introduction to PoolDilutionR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro_PoolDilutionR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is Pool Dilution?

Pool dilution is an isotope tracer technique wherein a biogeochemical pool is artifically enriched with its heavy isolog. After enrichment, the gross production and consumption rates of that pool can be calculated using the change in isotopic signature and absolute pool size over time. This technique can be applied to many different chemical species and was originally developed to measure soil nutrient transformations (Kirkham and Bartholomew 1954).  \


```{r setup, include=FALSE}
library(PoolDilutionR)
library(dplyr)
library(ggplot2)
library(ggpmisc)
library(tidyr)
library(kableExtra)

theme_set(theme_minimal() + theme(text = element_text(size = 11)))
```

## PoolDilutionR (PDR)

PDR uses equations originally published in von Fischer and Hedin 2002, 10.1029/2001GB001448, with some modifications. Unlike other pool dilution methods, this set of equations takes into account the effect of isotopic fractionation (see Fractionation below).  
```{r Input Data, message = FALSE}
Morris2023 %>%
  filter(id == "71") %>%
  select(-"id") %>%
  kbl(caption = "Data for Sample #71")
```

```{r Example Data, echo = FALSE, message = TRUE, out.width = c("100%", "300px")}
facet_labels <- c(
    AP_obs = "Atom% Observed",
    cal12CH4ml = "12C Methane (mL)",
    cal13CH4ml = "13C Methane (mL)")

Morris2023 %>%
    mutate(sample_id = as.factor(id)) %>%
    pivot_longer(cols = c(cal12CH4ml, cal13CH4ml, AP_obs)) %>%
    ggplot(aes(time_days, value,
               group = sample_id, color = sample_id)) +
    scale_color_discrete("Sample ID") +
    geom_point() + geom_line() + ylab("") + xlab("\n Timestep (d)") +
    ggtitle("Example Input Data") +
    facet_wrap(~name, scales = "free_y",
                labeller = labeller(name = facet_labels)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.75)) -> input_data
print(input_data)
```

## Fractionation
